---
title: "RepData_PeerAssessment2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(lubridate)
library(ggplot2)
```

## Title


## Synopsis


## Data Processing

```{r DataLoading, cache= TRUE}

# Downloading the data

#download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "./Data.csv.bz2", method = "curl")

Data <- read.csv("./Data.csv", header = T)
# Dim 902297*37

str(Data)

```

### Looking at the data

First we can aim to reducing the size of data. Filtering out unwanted data varaibles can be done in multiple ways like :

#### One can prioritize the variables required based on the question asked.

In our case our 2 important questions are :  
        1. Population Health  
        2. Economic loss  

Thus the columns like EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG seem relevant.  

```{r Subsetting_Relevant_Vars}
Imp_Vars <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG" )
Data.subset <- Data[,Imp_Vars]

```

#### We can also focus on the most frequent events as they're more likely to happen.  

Then first we see the frequency of each event and select the top fifty.  


```{r FrequentEvents}

# Limiting ourselves top 50 most frequent events.
Events <- table(Data.subset$EVTYPE)
Events <- sort(Events, decreasing = T)
#sum(Events[1:50]) == 894432; Thus our top 50 frequent incidents cover 99% data points/ rows. Rest of the 935 events are once in chance events probably.

Most_Freq_Events <- names(Events)[1:50]

Get_RowIndices_Event <- function(Event) { which(Data.subset$EVTYPE == Event)}

Row_I <- unlist(lapply(Most_Freq_Events, Get_RowIndices_Event))
Data.subset <- Data.subset[Row_I,]
# Dim 894432*6

```


#### Take an estimate of the missing data in the above shortlisted variables  

```{r MissingData}

FatalitybyEvent <- aggregate(FATALITIES + INJURIES~ EVTYPE, data = Data.subset, sum)

Data.subset$Year <-year(mdy_hms(Data.subset$BGN_DATE))
Newdf <- melt(Data.subset, id =c("EVTYPE", "Year"), measure.vars = 3:6 )

Year_Casualty <- dcast(Newdf, EVTYPE + Year ~ variable, sum)
'
ggplot( data = Year_Casualty, aes(x =Year)) + geom_line(aes(y = log2(INJURIES)), color = "firebrick1",size = 1) +   geom_line(aes(y = log2(FATALITIES)), color = "darkred", linetype="twodash") + 
geom_line(aes(y = log2(PROPDMG)), color = "darkslateblue") +
geom_line(aes(y = log2(CROPDMG)), color = "darkolivegreen") +
    facet_wrap(~EVTYPE)'

Event_Casualty <- dcast(Newdf,EVTYPE ~ variable, sum)

set.seed(42)
require("ggrepel")
'
ggplot( data = Event_Casualty, aes( x = log10(FATALITIES),y=log10(INJURIES))) + 
        geom_point(color=1:50) +  geom_text_repel(aes(label= EVTYPE),  size=5)
        

ggplot( data = Event_Casualty, aes( x = log10(PROPDMG),y=log10(CROPDMG))) + 
        geom_point(color=1:50) +  geom_text_repel(aes(label= EVTYPE),  size=5)
'
```







